---
title: "Basecalling function for Illumina BCL"
author: "Thomas Carroll and Marian Dore"
date: "`r doc_date()`"
package: "`r pkg_ver('basecallQC')`"
abstract: >
  Instructions on enabling _Bioconductor_ style 
  in your _R_ markdown vignettes.
vignette: >
  %\VignetteIndexEntry{Bioconductor style for HTML documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document
---

# Prerequisites

In order to build .Rmd documents the `r CRANpkg("rmarkdown")` package is necessary, which relies on [pandoc](http://johnmacfarlane.net/pandoc/) to compile the final output document. If you are working within a recent version of RStudio then both rmarkdown and pandoc should be already installed.

If you are missing the `r CRANpkg("rmarkdown")` package then you can install it from CRAN using

```{r rmarkdown, eval=FALSE}
install.packages("rmarkdown")
```

In this case a recent version of [pandoc](http://johnmacfarlane.net/pandoc/) (>= 1.12.3) is also required. See the [pandoc installation instructions](https://github.com/rstudio/rmarkdown/blob/master/PANDOC.md) for details on installing pandoc for your platform.


# Authoring R markdown HTML package vignettes

```{r code, echo = FALSE}
library(basecallQC)
fileLocations <- system.file("extdata",package="basecallQC")

demuxStats <- dir(fileLocations,pattern="DemultiplexingStats.xml",full.names=TRUE)
p <- processDemultiplex(demuxStats)
```



# Session info

Here is the output of `sessionInfo()` on the system on which this
document was compiled:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```


[R]: http://r-project.org
[RStudio]: http://www.rstudio.com/
