---
title: "Working with Illumina basecalling data using the basecallQC package"
author: "Marian Dore and Thomas Carroll"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## The basecallQC package

The basecallQC package provides a set of tools to streamline basecalling and demultiplexing using the Illumina bcl2Fastq software (versions > 2.1.8).

The basecallQC package includes functions too:- 

- "Clean" Illumina bcl2Fastq (versions > 2.1.8) samplesheets e.g. Remove invalid sample names, whitespaces, correct column headers
- Convert Samplesheets from Illumina bcl2Fastq versions < 1.8.9 to be compatable to Illumina bcl2Fastq versions > 2.1.8.
- Create base masks to used for Illumina basecalling with bcl2Fastq versions > 2.1.8.
- Create summary tables and plots from Illumina basecalling with bcl2Fastq versions > 2.1.8.


## Cleaning and updating the Sample Sheet

The Illumina bcl2Fastq program requires a sample sheet to dictate sample names, indexes and parameters for basecalling and demultiplexing. 

The construction of sample sheets for basecalling and demultiplexing requires specific column names,valid names Sample IDs and correct indexes. The complexity of working with Illumina sample sheets is compounded by the use of different sample sheet specifications before and after version 1.8.9 of the bcl2Fastq program.

The basecallQC package uses the **validateBCLSheet** function to both "clean" sample sheets to have valid column and sample names as well as to update sample sheets for versions < 1.8.9 to be compatable with version > 2.18   

To update a samplesheet, first the parameters for the Run must be captured using the **setBCL2FastQparams** command. The resulting  BCL2FastQparams object contains information on the Run parsed from the RunParameters.xml such as the actual index length which may be required to correct the sample sheet. 

Once the BCL2FastQparams object has been created, a data.frame or matrix can be passed to the **validateBCLSheet** alongside the BCL2FastQparams object to create a valid sample sheet object,  which may then be written to file.

### Cleaning a sample sheet.

In this first example we pass the **validateBCLSheet** function a sample sheets for versions < 1.8.9 and update it to be compatable with version > 2.18

```{r, cleaning, fig.show='hold',warning=FALSE,message=FALSE}
library(basecallQC)
fileLocations <- system.file("extdata",package="basecallQC")
config <- dir(fileLocations,pattern="config.ini",full.names=TRUE)
fileLocations <- system.file(file.path("extdata","testSampleSheets"),package="basecallQC")
runXML <- dir(fileLocations,pattern="runParameters.xml",full.names=TRUE)
sampleSheet <- dir(fileLocations,pattern="*\\.csv",full.names=TRUE)
read.delim(sampleSheet[1],sep=",",header = T,comment.char = "[")

```

This sample sheet contains invalid Sample names e.g. A 1 (whitespace) and 08.HDOX37 (starts with numeric) as well as invalid column headers such as "Name". 

```{r, cleaning2, fig.show='hold',warning=FALSE,message=FALSE}

bcl2fastqparams <- setBCL2FastQparams(runXML,config,runDir=getwd(),verbose=FALSE)
cleanedSampleSheet <- validateBCLSheet(sampleSheet[1],param=bcl2fastqparams)
head(cleanedSampleSheet)

```

In the cleaned samplesheet the column headers have been corrected and invalid Sample IDs converted to valid IDs e.g.  08.HDOX37 has been converted to Sample_HDOX37

### Updating a sample sheet.

Updating a sample sheet from those used with versions <= 1.8.9 to one compatable with versions >= 2.1.8 is done following the same procedure as cleaning a sample sheet.


```{r, updating, fig.show='hold',warning=FALSE,message=FALSE}
library(basecallQC)
fileLocations <- system.file("extdata",package="basecallQC")
config <- dir(fileLocations,pattern="config.ini",full.names=TRUE)
fileLocations <- system.file(file.path("extdata","testSampleSheets"),package="basecallQC")
runXML <- dir(fileLocations,pattern="runParameters.xml",full.names=TRUE)
sampleSheet <- dir(fileLocations,pattern="*\\.csv",full.names=TRUE)
read.delim(sampleSheet[2],sep=",",header = T,comment.char = "[")

```

The sample sheet from versions <= 1.8.9 contains many header which will need to be updated to those used versions >= 2.1.8 as well as superflous header which are maintained as metadata.

The specification of indexes in sample sheets from versions <= 1.8.9 were in a single column with indexes separated by a hyphon. This specification has changed to the explicit inclusion of an index2 column when using dual indexes so this to must be udpated.

```{r, updating2, fig.show='hold',warning=FALSE,message=FALSE}

bcl2fastqparams <- setBCL2FastQparams(runXML,config,runDir=getwd(),verbose=FALSE)
cleanedSampleSheet <- validateBCLSheet(sampleSheet[2],param=bcl2fastqparams)
head(cleanedSampleSheet)

```

In the resulting updated sample sheet we can see column names have been updated,  "SampleID" to "Sample_ID" and the index column has been automatically split into index and index2 columns.

## Base masking and the bcl2Fastq2 command.

The basecallQC package can also provide the base masks and basecalling command for bcl2Fastq2 versions > 2.1.8 from the cleaned sample sheet and the BCL2FastQparams object.

The **createBaseMasks** function creates a tibble data.frame of basemasks per lane from a cleaned sample sheet, as from validateBCLsheet, and a BCL2FastQparams object.

```{r, basemasks, fig.show='hold',warning=FALSE,message=FALSE}
fileLocations <- system.file("extdata",package="basecallQC")
runXML <- dir(fileLocations,pattern="runParameters.xml",full.names=TRUE)
config <- dir(fileLocations,pattern="config.ini",full.names=TRUE)
sampleSheet <- dir(fileLocations,pattern="*\\.csv",full.names=TRUE)
bcl2fastqparams <- setBCL2FastQparams(runXML,config,runDir=getwd(),verbose=FALSE)
cleanedSampleSheet <- validateBCLSheet(sampleSheet,param=bcl2fastqparams)
baseMasks <- createBasemasks(cleanedSampleSheet,param=bcl2fastqparams)
baseMasks$index1Mask
```

Following the creation of a base masks tibble data.frame, the **createBCLcommand** function takes the cleaned sample sheet, a BCL2FastQparams object and the base masks data.frame to create the command to be used for Illumina basecalling/demultiplexing using  bcl2Fastq2 versions > 2.1.8.

The command is returned as a character string to allow the user to control submission of command to best fit the user's system.

```{r, submitCommand, fig.show='hold',warning=FALSE,message=FALSE}
toSubmit <- createBCLcommand(bcl2fastqparams,cleanedSampleSheet,baseMasks)
toSubmit
```


## Running the basecallQC pipeline

The basecallQC package provides a pipeline function to allow the user to clean/convert the samplesheet, create the bcl2Fastq2 command and parse any basecalling/demultiplexing results. The resulting basecallQC object can then be used to produce summary tables, plots and a report.

The basecallQC function takes a BCL2FastQparams object for the Run, a cleaned sample sheet and any Run metadata that the user wishes to attach to their experiment and returns a **basecallQC** object.

The resulting **basecallQC** object can be used with basecallQC's plotting and reporting functions.

```{r, basecallQC, fig.show='hold',warning=FALSE,message=FALSE}
fileLocations <- system.file("extdata",package="basecallQC")
runXML <- dir(fileLocations,pattern="runParameters.xml",full.names=TRUE)
config <- dir(fileLocations,pattern="config.ini",full.names=TRUE)
sampleSheet <- dir(fileLocations,pattern="*\\.csv",full.names=TRUE)
outDir <- file.path(fileLocations,"Runs/161105_D00467_0205_AC9L0AANXX/C9L0AANXX/")
bcl2fastqparams <- setBCL2FastQparams(runXML,config,runDir=getwd(),outDir,verbose=FALSE)
bclQC <- basecallQC(bcl2fastqparams,RunMetaData=NULL,sampleSheet)
class(bclQC)
```

## Plots, tables and reports from the basecallQC pipeline. 

The **basecallQC** object contains information on demultiplexing, basecalling and run parameters.

Summary HTML tables for demultiplexing and basecalling statitistics can be generated using the **summaryDemuxTable** and **summaryConvStatsTable** repsectively. The output argument can be set to "static" or "html" to allow for tables for use in non-interactive and interactive modes.


```{r, tables, fig.show='hold',warning=FALSE,message=FALSE}
summaryConvStatsTable(bclQC,output = "html")
summaryDemuxTable(bclQC,output = "html")
```

The user may also visualise the results using basecallQC's plotting functions for basecalling and demultiplexing stats.

The **passFilterBar** function produces a bar plot of yields per sample. The metric to plot and how to summarise data for plotting are controlled by the *metricToPlot* and *groupBy* arguments.

```{r, plots1, fig.show='hold',warning=FALSE,message=FALSE, fig.width=5, fig.height=5}
passFilterBar(bclQC,groupBy="Sample",metricToPlot = "Yield")

```

The **passFilterTilePlot** produces a plot of the specified metric across the Lanes and Tiles of the investigated Run.

```{r, plots2, fig.show='hold',warning=FALSE,message=FALSE, fig.width=5, fig.height=5}

passFilterTilePlot(bclQC,metricToPlot = "Yield")

```

The **demuxBarplot** function produces a bar plot of yield from demultiplexing. For demultiplexing statistics only the *groupBy* arguments cna be used.

```{r, plots3, fig.show='hold',warning=FALSE,message=FALSE, fig.width=5, fig.height=5}

demuxBarplot(bclQC,groupBy="Sample")
```

### Producing an HTML report.

The basecallQC package allows the user to generate a report per Illumina Run containing the most important demultiplexing and basecalling metrics. 

This report may be customised and the resulting RMD passed to the reportBCL function's *reportRMDfile* argument. This RMD will then be used in place of the standard basecallQC report.


```{r, plots4, eval=F, warning=FALSE,message=FALSE}
reportBCL(bclQC)
```


```{r sessionInfo,echo=F,fig.height=30,fig.width=15}
sessionInfo()
```
